Evolution of ORM Systems: Formal Foundations

### 2013-01-08 doc/description.tex
updated orm mappings and app transformation moved into one section

\subsubsection{Application Transformation}

@@

\subsubsection{Mapping of Types}
The mapping of types assumes a bijection between application types and database types, otherwise these has to be additional information for type mapping. We focus on structural and data change therefore we simplify types and its mapping.

\begin{equation*}
\begin{gathered}
	\rho_{t} : AppType \rightarrow DbType  \\
 	\rho_{t}(APPSTRING) = DBSTRING 
\end{gathered}
\end{equation*}

@@

\begin{center}
$\rho_{c}: Class \times Database \rightarrow Database $ \\ 
$\rho_{c}(c,d) = table(name(c), primaryKey("Id"), \varnothing, \varnothing) $
\end{center}

@@

$
\rho_p : Property \times Database \rightarrow Database 
$

$\rho_p(p, d) = \begin{cases}
  \inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d) 
 \\ \wedge \not instanciated(owningClass(p))
 }{\begin{gathered}
	  column(name(p), \rho_t(type(p)), \varnothing )
\end{gathered}
 } \\ \\

  \inference{cardinality(p) = 1 \wedge mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}
 	 column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
\end{gathered}} 
\\ \\

 \inference{ cardinality(p) > 1 \wedge !\;mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL) \\
\end{gathered}}
\\ \\

 \inference{ cardinality(p) > 1 \wedge mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d) }{\begin{gathered}  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), NOTNULL), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL)
	 \end{gathered}}
 \end{cases}$

@@

$
\rho_a(a,d) = \begin{cases}
 \inference{ startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)}{ 
 \begin{gathered}
 foreignKeys(u) \\ \cup foreignKey(name(a), name(t),  \varnothing) 
 \end{gathered}
 }
  \\ \\
 \inference{ startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)}{\begin{gathered}  
foreignKeys(u) \\ \cup  foreignKey(name(a), name(t),  NOTNULL) \\
\end{gathered}
} \\ \\

  \inference{  startCardinality > 1 \\ \wedge \rho(owningClass(p)) \in tables(d) \wedge \rho(reference(a)) \in tables(d)}{
  \begin{gathered}  
 table(name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  \end{gathered}}  
 \end{cases}
$

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $newClass$, $addProperty$, $addAssociation$. These three transformations can be used to create a software from a scratch. The first of them has no impact on stored data, whereas the rest of them can in some cases.
\begin{align*}
\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
\Phi(s, addProperty(l, p)) = database(s) \bullet insertProp(l, p) \\
s \in Softwares, l \in Label, p \in Properties
\end{align*}
Adding a property into an application structure has to verify a name collision only:
\begin{align*}
addProp : Label \times Property \times Application \rightarrow Application \\
addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
l \in Label, p \in Properties, a \in Applications
\end{align*}
Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{align*}
insertProp(p, d) = \begin{cases}
\inference{!\:mandatory(p) \vee !\:instantiated(owningClass(p))}{addColumn(l, \rho(p,d), d)} 
\\ \\ 
\inference{ \wedge mandatory(p) \vee instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered} addColumn(l, \rho(p,d), d) \\ \bullet \forall r \in data(d), reference(r) = name(t) : \\ appendValue(name(t), [name(p), defaultValue(p)], d)
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
addTable(\rho(p,d), d) \bullet \forall r \in row(owningClass(p)) :\\ insertRow(row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]),d) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{align*}

\begin{align*}
\mathbf{addAsociation} : Label \times Association \times (Row \times Row) \times State \rightarrow State \\
\Psi(s, addAsociation(l, a, f_m)) = application(s) \circ addAssoc(l, a) \\
\Phi(s, addAsociation(l, a, f_m)) = database(s) \bullet insertAssoc(l, a, f_m)
\end{align*}
Adding an association into an application structure has to verify if there is no name collision and if there is the associated class present in the model.
\begin{align*}
addAssoc : Label \times Association \times Application \rightarrow Application \\
addAssoc(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(c) \cup s}\\
l \in Labels, s \in Associations, a \in Applications
\end{align*}
Adding an association into a database is complicated because of already existing instances. When an association between classes is added the instances of both classes has to be paired. This pairing is provided by the function $f_m$:
$$f_m : \rightarrow Instance \times Instance $$

@@

\begin{align*}
\mathbf{newClass} : Class \times State \rightarrow State \\
\Psi(s, newClass(c)) = application(s) \circ addClass(c) \\
\Phi(s, newClass(c)) = database(s)  \bullet addTable(\rho(c,d), d) \\ \bullet addColumn(name(c), \{\forall r \in singleProperties(c) | \rho(r,d)\}, d) \\ \bullet addTable(name(c), \{\forall r \in collectionProperties(c) | \rho(r,d)\}, d)  
\end{align*}
\begin{align*}
addClass: Class \times Application \rightarrow Application \\ 
addClass(c, a) = \inference{\forall s \in classes(a) : name(s) \neq name(c)}{classes(a) \cup c} \\ \\
c \in Classes, a \in Applications
\end{align*}
\paragraph{Add Class} The $insertClass$ operation changes the structure of the database schema only, because the $insertAssoc$ transformation needs a mapping function to work properly, which is not part of the $newClass$ definition. The situation reflect the modeling praxis when the structure of a class is defined and then it is associations are created.
%\begin{align*}
%insertClass : Class \times Database \rightarrow Database \\
%insertClass(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
%	%\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) \\ 
%	assocaiations(c) = \varnothing
%}{
%\begin{gathered}
% addTable(\rho(c, d),d) \\ \wedge t \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ 
%\end{gathered}
%}
%\end{align*}

\subsection{Software Decomposition}
\begin{align*}
RemoveClass: Label \times Software \rightarrow State  \\
\Psi RemoveClass(l, s) = applications(s) \circ remClass(l) \\
\Phi RemoveClass(l, s) = database(s) \bullet dropClass(l) \\
\\ l \in Label, s \in Software 
\\ \\
remClass(l, a) = \inference{\exists c \in classes(a) : name(c) = l \\
\wedge !\: instantiated(c) = \varnothing \wedge !\: associated(c)} {classes(a) \setminus c } \\
\end{align*}
\begin{align*}
dropClass(l,d) = \inference{ \exists t \in tables(d) : name(t) = l  \\ \wedge \not \exists r \in data(d) : owningTable(r) = l \\  \not \exists j \in tables(d) \wedge f \in foreignKeys(j) : reference(f) = l }{ tables(d) \setminus \rho(c)}
\end{align*}

\begin{align*}
RemoveProperty: Label \times Label \times State \rightarrow State \\
\Psi RemoveProperty(l, l_p, s) =  applications(s) \circ  remProp(l, l_p) \\
\Phi RemoveProperty(l, l_p, s) =  database(s) \bullet dropProp(l, l_p) \\
RemoveProperty(l_c, l_p, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{properties(c) \setminus p }\\
l_c, l_p \in Label, a \in Applications 
\end{align*}
\begin{align*}
dropProp(p, d) = \begin{cases}
 \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \exists c \in columns(t) : name(c) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p)) }{columns(t) \setminus c
} \\ \\
  \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \not \exists c \in columns(t) : name(c) = name(p) \\ \wedge \exists u \in tables(d) : name(u) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p))}{ tables(d) \setminus u }
 \end{cases}
\end{align*}

\begin{align*}
RemoveAssociation : Label \times Labels \times State \rightarrow State \\ 
\Psi RemoveAssociation(l, l_a, s) = application(s) \circ remAss(l, l_a) \\
\Phi RemoveAssociation(l, l_a, s) = database(s) \bullet dropAssoc(l, l_a) \\
RemoveAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists f \in associations(c) : name(f) = l_a }{associations(c) \setminus f} \\
l_c, l_a \in Labels, a \in Applications
\end{align*}
\begin{align*}
dropAssoc(l, l_a, d) =  \begin{cases}
 \inference{\exists t \in tables(d) : name(t) = l
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = l_a
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ foreignKeys(t) \setminus f }
 \\ \\
 \inference{\exists t \in tables(d) : name(t) = l_a
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = l
 \\ \wedge \forall r \in data(d) : reference(r) \neq t}{ tables(d) \setminus t}
 \end{cases}
\end{align*}

\subsection{Software Alternation}
\paragraph{Rename Class}
\begin{align*}
RenameClass : Label \times Label \times State \rightarrow State \\ 
\Psi(RenameClass(l_o, l_n), s) = application(s) \circ renClass(l_o, l_n) \\
\Phi(RenameClass(l_o, l_n), s) = database(s) \bullet alterClass(l_o, l_n) \\ \\
renClass:  Label \times Label \times Application \rightarrow Application \\
renClass(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_o \\ \wedge \forall e \in classes(a) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n \wedge \forall e \in classes(a) \\ \wedge \forall (f \in associations(e) \wedge reference(f) = l_o ) \\ : f.reference = l_n 
\end{gathered}}\\ \\
alterTable:  Label \times Label \times Database \rightarrow Database \\
alterTable(l_o, l_n, d) = \inference{\exists t \in tables(d) : name(d) = l_o \\ \wedge \forall e \in tables(d) : name(e) \neq l_n}{\begin{gathered}
t.name = l_n \\ \wedge \forall e \in tables(a) \wedge \forall (f \in fks(e) \wedge reference(f) = l_o ) \\ : f.reference = l_n 
\end{gathered}}\\
l_o, l_n \in Label, s \in States, a \in Applications, d \in Databases
\end{align*}

\paragraph{Rename Property}
\begin{align*}
RenameProperty : Label \times Label \times Label \times State \rightarrow State \\ \\
\Psi(RenameProperty(l_c, l_o, l_n, s)) = application(s) \circ renProp(l_c, l_o, l_n) \\
\Phi(RenameProperty(l_c, l_o, l_n, s)) = database(s) \bullet alterColumn(l_c, l_o, l_n) \\\\
renProp : Label \times Label \times Label \times Application \rightarrow Application \\
	renProp(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n}{\begin{gathered}
p.name = l_n 
\end{gathered}}\\\\
alterColumn : Label \times Label \times Label \times Database \rightarrow Database \\
	alterColumn(l_c, l_o, l_n, d) = \begin{cases}
 \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_o \\ \wedge \forall e \in columns(t) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n 
\end{gathered}}
\\ \\
\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
t.name = l_n 
\end{gathered}}
 \end{cases}
 \\
l_c, l_o, l_n \in Label, a \in Applications, s \in States, d \in Databases
\end{align*}

\paragraph{Rename Association}
\begin{align*}
RenameAssociation : Label \times Label \times Label \times State \rightarrow State \\
\Psi(RenameAssocaiation(l_c, l_o, l_n), s) = application(s) \circ renAssoc(l_c, l_o, l_n)\\
\Phi(RenameAssocaiation(l_c, l_o, l_n), s) = \begin{cases}
	\inference{\mathbf{TODO}}{database(s) \bullet alterTable(l_c, l_o, l_n)}
 \\
  \inference{\mathbf{TODO}}{database(s) \bullet alterColumn(l_c, l_o, l_n)}
 \end{cases}
\\\\
renAssoc(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists b \in associations(c) : name(b) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n}{\begin{gathered}
b.name = l_n 
\end{gathered}}\\\\
alterColumn : Label \times Label \times Label \times Database \rightarrow Database \\
%	alterAssoc(l_c, l_o, l_n, d) = \begin{cases}
%?? \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists f \in fks(t) : name(c) = l_o \\ \wedge \forall e \in fks(t) : name(e) \neq l_n}{\begin{gathered}
%f.name = l_n 
%\end{gathered}}
%\\ \\
%\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
%t.name = l_n 
%\end{gathered}}
% \end{cases}
% \\
l_c, l_o, l_n \in Label, a \in Applications, s \in Sates, d \in Databases
\end{align*}

>>>>

@@

\subsubsection{Mapping of Types}
The mapping of types assumes a bijection between application types and database types, otherwise these has to be additional information for type mapping. We focus on structural and data change therefore we simplify types and its mapping.
\begin{align*}
	\mathbf{\rho_{t}} : AppType \rightarrow DbType  \\
 	\rho_{t}(APPSTRING) = DBSTRING 
\end{align*}

@@

\begin{align*}
	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
	\rho_{c}(c) = \begin{cases}
		\inference{c = \varnothing}{\varnothing}\\\\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \\ \\
		\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))}
 \end{cases}\\
 c = Class
\end{align*}

@@

\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
 	\end{cases}
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
 	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
 		\wedge 	\exists t : t = table(name(owningClass(p)), d)
 		}{\begin{gathered}
 			table(name(p), primaryKey("Id"), \\ column("value", 			\rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
	\end{gathered}}\\ \\
 	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
 		\wedge \exists t : t = table(name(owningClass(p)), d) 
 		}{\begin{gathered}  
 			table(name(p), primaryKey("Id"), \\ 
 			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
	 \end{gathered}}
 	\end{cases}
\end{align*}

@@

\begin{align*}
	\mathbf{\rho_a} : Association* \rightarrow ForeignKey* \\
	\rho_a(a) = \begin{cases}
		\inference{ startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
		}{
			foreignKey(name(a), name(t),  \varnothing) 
	 	}
  \\ \\
 	 \inference{ startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
 	 }{ 
		foreignKey(name(a), name(t),  NOTNULL)
	}
	 \end{cases}
\end{align*}
\begin{align*} 
	\mathbf{\rho_a} : Association* \rightarrow Table* \\
  	\rho_a(a) = \inference{  startCardinality > 1 \\ \wedge \rho(owningClass(p)) \in tables(d) \wedge \rho(reference(a)) \in tables(d)
  		}{\begin{gathered}  
		 table(name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  		\end{gathered}}  
\end{align*}

\section{Model Manipulating Transformation}
\subsection{Application Transformation}
\paragraph{Add Property}Adding a property into an application structure has to verify a name collision only:
\begin{align*}
 	\mathbf{addProp} : Label \times Property \times Application \rightarrow Application \\
	addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
	l \in Label, p \in Properties, a \in Applications
\end{align*}
\paragraph{Add Association} Adding an association into an application structure has to verify if there is no name collision and if there is the associated class present in the model.
\begin{align*}
	\mathbf{addAssoc} : Label \times Association \times Application \rightarrow Application \\
	addAssoc(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(c) \cup s}\\
	l \in Labels, s \in Associations, a \in Applications
\end{align*}
\paragraph{Add Class} Adding a class is limited to the case when the class has no associations, this is because it simplifies the database migration and because it is a natural way of modeling. Next the transformation $addClass$ has to verify there is no name collision
\begin{align*}
	\mathbf{addClass} : Class \times Application \rightarrow Application \\ 
	addClass(c, a) = \inference{\forall s \in classes(a) : name(s) \neq name(c) \\ \wedge associations(c) = \varnothing }{classes(a) \cup c} \\ \\
	c \in Classes, a \in Applications
\end{align*}
\paragraph{Remove Property}
\begin{align*}
 	\mathbf{remProp} : Label \times Label \times Application \rightarrow Application \\
 	remProp(l_c, l_p, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p
	}{properties(c) \setminus p }\\
	l_c, l_p \in Label, a \in Applications 
\end{align*}
\paragraph{Remove Association}
\begin{align*}
	\mathbf{remAssoc} : Label \times Label \times Application \rightarrow Application \\
	remAssoc(l_c, l_a, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists f \in associations(c) : name(f) = l_a }{associations(c) \setminus f} \\
	l_c, l_a \in Labels, a \in Applications
\end{align*}
\paragraph{Remove Class} The class can be removed only in case it is not referenced by any association existing in the application.
\begin{align*}
	\mathbf{remClass} : Label \times Application \rightarrow Application \\
	remClass(l, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge !\: associated(c)
	} {classes(a) \setminus c }\\
	l \in Label, a = Application
\end{align*}
\paragraph{Rename Property}
\begin{align*}
	\mathbf{renProp} : Label \times Label \times Label \times Application \rightarrow Application \\
	renProp(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n
	}{\begin{gathered}
		p.name = l_n 
	\end{gathered}}
\end{align*}
\paragraph{Rename Association}
\begin{align*}
	\mathbf{renAssoc} : Label \times Label \times Label \times Application \rightarrow Application \\
	renAssoc(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists b \in associations(c) : name(b) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n
	}{\begin{gathered}
		b.name = l_n 
	\end{gathered}}
\end{align*}
\paragraph{Rename Class}
\begin{align*}
	\mathbf{renClass} :  Label \times Label \times Application \rightarrow Application \\
	renClass(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_o \\ \wedge \forall e \in classes(a) : name(e) \neq l_n
	}{\begin{gathered}
		c.name = l_n \wedge \forall e \in classes(a) \\ \wedge \forall (f \in associations(e) \wedge reference(f) = l_o ) \\ : f.reference = l_n 
	\end{gathered}}
\end{align*}

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $newClass$, $addProperty$, $addAssociation$. These three transformations can be used to create a software from a scratch. The first of them has no impact on stored data, whereas the rest of them can in some cases.
\begin{align*}
\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
\Phi(s, addProperty(l, p)) = database(s) \bullet insertProp(l, p) \\
s \in Softwares, l \in Label, p \in Properties
\end{align*}

Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{align*}
insertProp(p, d) = \begin{cases}
\inference{!\:mandatory(p) \vee !\:instantiated(owningClass(p))}{addColumn(l, \rho(p,d), d)} 
\\ \\ 
\inference{ \wedge mandatory(p) \vee instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered} addColumn(l, \rho(p,d), d) \\ \bullet \forall r \in data(d), reference(r) = name(t) : \\ appendValue(name(t), [name(p), defaultValue(p)], d)
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
addTable(\rho(p,d), d) \bullet \forall r \in row(owningClass(p)) :\\ insertRow(row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]),d) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{align*}

\begin{align*}
\mathbf{addAsociation} : Label \times Association \times (Row \times Row) \times State \rightarrow State \\
\Psi(s, addAsociation(l, a, f_m)) = application(s) \circ addAssoc(l, a) \\
\Phi(s, addAsociation(l, a, f_m)) = database(s) \bullet insertAssoc(l, a, f_m)
\end{align*}

Adding an association into a database is complicated because of already existing instances. When an association between classes is added the instances of both classes has to be paired. This pairing is provided by the function $f_m$:
$$f_m : \rightarrow Instance \times Instance $$

@@

\begin{align*}
\mathbf{newClass} : Class \times State \rightarrow State \\
\Psi(s, newClass(c)) = application(s) \circ addClass(c) \\
\Phi(s, newClass(c)) = database(s)  \bullet addTable(\rho(c,d), d) \\ \bullet addColumn(name(c), \{\forall r \in singleProperties(c) | \rho(r,d)\}, d) \\ \bullet addTable(name(c), \{\forall r \in collectionProperties(c) | \rho(r,d)\}, d)  
\end{align*}

\paragraph{Add Class} The $insertClass$ operation changes the structure of the database schema only, because the $insertAssoc$ transformation needs a mapping function to work properly, which is not part of the $newClass$ definition. The situation reflect the modeling praxis when the structure of a class is defined and then it is associations are created.
%\begin{align*}
%insertClass : Class \times Database \rightarrow Database \\
%insertClass(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
%	%\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) \\ 
%	assocaiations(c) = \varnothing
%}{
%\begin{gathered}
% addTable(\rho(c, d),d) \\ \wedge t \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ 
%\end{gathered}
%}
%\end{align*}

\subsection{Software Decomposition}
\begin{align*}
RemoveClass: Label \times Software \rightarrow State  \\
\Psi RemoveClass(l, s) = applications(s) \circ remClass(l) \\
\Phi RemoveClass(l, s) = database(s) \bullet dropClass(l) \\
\\ l \in Label, s \in Software 
\end{align*}
\begin{align*}
dropClass(l,d) = \inference{ \exists t \in tables(d) : name(t) = l  \\ \wedge \not \exists r \in data(d) : owningTable(r) = l \\  \not \exists j \in tables(d) \wedge f \in foreignKeys(j) : reference(f) = l }{ tables(d) \setminus \rho(c)}
\end{align*}

\begin{align*}
RemoveProperty: Label \times Label \times State \rightarrow State \\
\Psi RemoveProperty(l, l_p, s) =  applications(s) \circ  remProp(l, l_p) \\
\Phi RemoveProperty(l, l_p, s) =  database(s) \bullet dropProp(l, l_p) \end{align*}
\begin{align*}
dropProp(p, d) = \begin{cases}
 \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \exists c \in columns(t) : name(c) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p)) }{columns(t) \setminus c
} \\ \\
  \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \not \exists c \in columns(t) : name(c) = name(p) \\ \wedge \exists u \in tables(d) : name(u) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p))}{ tables(d) \setminus u }
 \end{cases}
\end{align*}

\begin{align*}
RemoveAssociation : Label \times Labels \times State \rightarrow State \\ 
\Psi RemoveAssociation(l, l_a, s) = application(s) \circ remAss(l, l_a) \\
\Phi RemoveAssociation(l, l_a, s) = database(s) \bullet dropAssoc(l, l_a) \\
l_c, l_a \in Labels, a \in Applications
\end{align*}
\begin{align*}
dropAssoc(l, l_a, d) =  \begin{cases}
 \inference{\exists t \in tables(d) : name(t) = l
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = l_a
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ foreignKeys(t) \setminus f }
 \\ \\
 \inference{\exists t \in tables(d) : name(t) = l_a
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = l
 \\ \wedge \forall r \in data(d) : reference(r) \neq t}{ tables(d) \setminus t}
 \end{cases}
\end{align*}

\subsection{Software Alternation}
\paragraph{Rename Class}
\begin{align*}
RenameClass : Label \times Label \times State \rightarrow State \\ 
\Psi(RenameClass(l_o, l_n), s) = application(s) \circ renClass(l_o, l_n) \\
\Phi(RenameClass(l_o, l_n), s) = database(s) \bullet alterClass(l_o, l_n) \\ \\
\end{align*}

\begin{align*}
alterTable:  Label \times Label \times Database \rightarrow Database \\
alterTable(l_o, l_n, d) = \inference{\exists t \in tables(d) : name(d) = l_o \\ \wedge \forall e \in tables(d) : name(e) \neq l_n}{\begin{gathered}
t.name = l_n \\ \wedge \forall e \in tables(a) \wedge \forall (f \in fks(e) \wedge reference(f) = l_o ) \\ : f.reference = l_n 
\end{gathered}}\\
l_o, l_n \in Label, s \in States, a \in Applications, d \in Databases
\end{align*}

\paragraph{Rename Property}
\begin{align*}
	RenameProperty : Label \times Label \times Label \times State \rightarrow State \\ \\
\Psi(RenameProperty(l_c, l_o, l_n, s)) = application(s) \circ renProp(l_c, l_o, l_n) \\
\Phi(RenameProperty(l_c, l_o, l_n, s)) = database(s) \bullet alterColumn(l_c, l_o, l_n) \\\\
\end{align*}

\begin{align*}
alterColumn : Label \times Label \times Label \times Database \rightarrow Database \\
	alterColumn(l_c, l_o, l_n, d) = \begin{cases}
 \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_o \\ \wedge \forall e \in columns(t) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n 
\end{gathered}}
\\ \\
\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
t.name = l_n 
\end{gathered}}
 \end{cases}
 \\
l_c, l_o, l_n \in Label, a \in Applications, s \in States, d \in Databases
\end{align*}

\paragraph{Rename Association}
\begin{align*}
RenameAssociation : Label \times Label \times Label \times State \rightarrow State \\
\Psi(RenameAssocaiation(l_c, l_o, l_n), s) = application(s) \circ renAssoc(l_c, l_o, l_n)\\
\Phi(RenameAssocaiation(l_c, l_o, l_n), s) = \begin{cases}
	\inference{\mathbf{TODO}}{database(s) \bullet alterTable(l_c, l_o, l_n)}
 \\
  \inference{\mathbf{TODO}}{database(s) \bullet alterColumn(l_c, l_o, l_n)}
 \end{cases}
\\\\
alterColumn : Label \times Label \times Label \times Database \rightarrow Database \\
%	alterAssoc(l_c, l_o, l_n, d) = \begin{cases}
%?? \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists f \in fks(t) : name(c) = l_o \\ \wedge \forall e \in fks(t) : name(e) \neq l_n}{\begin{gathered}
%f.name = l_n 
%\end{gathered}}
%\\ \\
%\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
%t.name = l_n 
%\end{gathered}}
% \end{cases}
% \\
l_c, l_o, l_n \in Label, a \in Applications, s \in Sates, d \in Databases
\end{align*}
